---
title: "Simulated Causal Analysis of PFAS Effects on Infant Neurodevelopment via Placental Imprinting"
author: "Sean T. Bresnahan"
date: "2025-07-25"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

This script simulates a causal mediation framework in which maternal PFAS exposure influences infant neurodevelopment (SRS scores) via changes in placental parent-of-origin isoform expression. The mediation pathway is moderated by fetal sex, and maternal DNA methylation is leveraged as a negative control to adjust for shared genetic and environmental confounding. I compare three methods: Control Outcome Calibration (COCA), Proximal g-Computation (PGC), and regression adjustment models that include the control outcome variable as a covariate as recommended by Huang (2021).

#### Dataset:

- 100 infants
- 100 isoforms (each with 1–5 CpG sites)
- Maternal PFAS exposure (log-normal)
- Fetal sex (binary)
- Latent shared confounder (correlated with PFAS)
- Isoform-level maternal expression ratios (simulating imprinting)
- Infant SRS scores (influenced by disrupted imprinting and confounder)
- PFAS affects imprinting of a subset of isoforms
- Some imprinting–SRS associations are confounded by maternal methylation

# Load Packages and Define Knitr Options

```{r data_setup, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(lavaan)
library(boot)
library(ggplot2)
library(knitr)
```

# Simulate Dataset

We first define the number of samples (infants) and isoforms to be simulated.

## Define Samples and Isoforms

```{r parameters}
set.seed(42)
n_samples <- 100
n_isoforms <- 100
simulation_scenario <- "null"
```

Here we simulate fetal sex, maternal PFAS exposure levels, and a latent confounder that is correlated with PFAS.

## Generate Fetal Sex, PFAS Exposure, and Shared Confounder

```{r simulate_covariates}
# Fetal sex (0 = female, 1 = male)
fetal_sex <- rbinom(n_samples, size = 1, prob = 0.5)
hist(fetal_sex, main = "Fetal Sex", xlab = "Sex (0 = Female, 1 = Male)")

# Maternal PFAS exposure (ng/mL, log-normal)
# Common sdlog values for environmental contaminants in humans can range from 0.8 to 1.5.
PFAS <- rlnorm(n_samples, meanlog = log(3), sdlog = 0.9) # Adjusted sdlog from 0.4 to 0.9
hist(PFAS, main = "PFAS Exposure", xlab = "ng/mL")

# Latent confounder U, correlated with PFAS
U <- rnorm(n_samples, mean = scale(PFAS), sd = 1)
```

Next, we simulate maternal DNA methylation data for each isoform, capturing shared environmental or genetic confounding via a latent variable.

## Simulate CpG Methylation from Shared Confounding Signal

```{r simulate_methylation}
n_cpg_per_isoform <- sample(1:5, size = n_isoforms, replace = TRUE)
maternal_meth_list <- lapply(n_cpg_per_isoform, function(n_cpg) {
  matrix(
    plogis(matrix(rnorm(n_samples * n_cpg, mean = rep(U, each = n_cpg), sd = 0.5), nrow = n_samples)), # Adjusted sd from 1 to 0.5
    nrow = n_samples
  )
})
```

## Simulate Isoform-Level Expression Ratios

```{r simulate_expression_ratios}
# Start with baseline biallelic expression + noise (maternal expression ratio)
# Using Beta distribution for ratios (bounded 0-1), which is more appropriate for proportions.
# shape1 = 50, shape2 = 50 gives a mean of 0.5 and a reasonable, but not extremely narrow, variance.
expr_ratios <- matrix(rbeta(n_samples * n_isoforms, shape1 = 50, shape2 = 50), nrow = n_samples)
hist(expr_ratios, main = "Baseline parental-origin expression ratios", xlab = "Ratio M:M+P")

# Define 50% of isoforms as true mediators
mediated_isoforms <- sample(1:n_isoforms, size = n_isoforms * 0.5)

# PFAS influences imprinting in these isoforms
for (i in mediated_isoforms) {
  effect_mean <- 0.05 # Original mean effect
  if (simulation_scenario == "null") {
    effect_mean <- 0 # Set mean effect to 0 for null scenario
  }
  effect <- scale(PFAS) * rnorm(1, mean = effect_mean, sd = 0.01) # Use the adjusted mean
  expr_ratios[, i] <- expr_ratios[, i] + effect

  # Clamp values to ensure they stay within [0,1]
  expr_ratios[, i] <- pmax(0, pmin(1, expr_ratios[, i])) 
}

hist(expr_ratios, main = "Parental-origin expression ratios with PFAS association", xlab = "Ratio M:M+P")
```

## Simulate SRS Scores Influenced by Disrupted Imprinting

```{r simulate_srs}
# Create SRS score from expression of mediated isoforms + confounder + noise
srs_effect_mean <- 2 # Original mean effect
if (simulation_scenario == "null") {
  srs_effect_mean <- 0 # Set mean effect to 0 for null scenario
}
SRS_contrib <- rowSums(scale(expr_ratios[, mediated_isoforms]) %*%
                        matrix(rnorm(length(mediated_isoforms), mean = srs_effect_mean, sd = 0.5), ncol = 1)) # Use the adjusted mean
SRS <- scale(SRS_contrib + scale(U) * 0.5 + rnorm(n_samples, 0, 1))

hist(SRS, main = "Simulated SRS", xlab = "SRS")
```

## Include Confounding Effects

```{r include_confounding}
# Select 50% of mediated isoforms to be confounded by methylation
confounded_idx <- sample(mediated_isoforms, size = length(mediated_isoforms) * 0.5)
confounding_contrib <- rep(0, n_samples)

for (i in confounded_idx) {
  meth_mean <- rowMeans(maternal_meth_list[[i]])
  effect_meth_expr <- scale(meth_mean) * rnorm(1, mean = 0.2, sd = 0.05)
  expr_ratios[, i] <- expr_ratios[, i] + effect_meth_expr
  expr_ratios[, i] <- pmax(0, pmin(1, expr_ratios[, i]))
  
  confounding_contrib <- confounding_contrib + scale(meth_mean) * rnorm(1, mean = 1.5, sd = 0.3)
}

# Inject confounding into SRS
SRS <- scale(SRS + confounding_contrib)
hist(SRS, main = "Simulated SRS with methylation-driven bias", xlab = "SRS")
```

Finally, we clean up and scale the simulated matrices, then reshape the methylation data for downstream analysis and visualization.

## Finalize and Format Output

```{r finalize_output}
expr_df <- as.data.frame(expr_ratios)
colnames(expr_df) <- paste0("isoform_", 1:n_isoforms)

metadata_df <- data.frame(
  sample = 1:n_samples,
  SRS = SRS,
  PFAS = PFAS,
  fetal_sex = fetal_sex
)

meth_long <- do.call(rbind, lapply(1:n_isoforms, function(i) {
  meth_mat <- maternal_meth_list[[i]]
  n_cpg <- ncol(meth_mat)
  data.frame(
    sample = rep(1:n_samples, times = n_cpg),
    isoform = paste0("isoform_", i),
    cpg = rep(1:n_cpg, each = n_samples),
    meth = as.vector(meth_mat)
  )
}))

hist(meth_long$meth, main = "Isoform-Level Methylation", xlab = "Beta-value")
```

# Moderated Mediation Analysis (lavaan)

Select one isoform for demonstration.

```{r moderated_mediation}
iso <- "isoform_27"
df <- cbind(metadata_df, expr = expr_df[[iso]])

meth_avg <- meth_long %>%
  filter(isoform == iso) %>%
  group_by(sample) %>%
  summarise(m_meth = mean(meth), .groups = "drop")

df <- merge(df, meth_avg, by = "sample")

model <- '
  expr ~ a1*PFAS + a2*fetal_sex + a3*PFAS:fetal_sex
  SRS ~ b1*expr + b2*fetal_sex + b3*expr:fetal_sex + c*PFAS
  ind_female := a1 * b1
  ind_male := (a1 + a3) * (b1 + b3)
  dir := c
  total_female := dir + ind_female
  total_male := dir + ind_male
'

fit <- sem(model, data = df, se = "bootstrap", bootstrap = 100)
summary(fit, standardized = TRUE, fit.measures = TRUE)
```

# COCA (Control Outcome Calibration)

```{r coca}
# COCA calibration model: includes SRS × fetal_sex interaction
coca_model <- lm(m_meth ~ PFAS + SRS * fetal_sex, data = df)

# Extract coefficients for calibration formula
beta_A <- coef(coca_model)["PFAS"]
beta_Y <- coef(coca_model)["SRS"]

# Calibrated PFAS effect estimate
psi_coca <- -beta_A / beta_Y
cat("COCA-calibrated PFAS effect estimate:", round(psi_coca, 4), "\n")

coca_fn <- function(data, indices) {
  d <- data[indices, ]
  fit <- lm(m_meth ~ PFAS + SRS * fetal_sex, data = d)
  -coef(fit)["PFAS"] / coef(fit)["SRS"]
}

boot_out <- boot(df, coca_fn, R = 100)
boot.ci(boot_out, type = "perc")
```

# Proximal G-Computation (PGC)

```{r pgc}
# Stage 1: bridge function includes interaction (optional but aligned with SEM)
bridge_model <- lm(m_meth ~ expr + SRS * fetal_sex, data = df)
df$Z_hat <- predict(bridge_model)

# Stage 2: PFAS × fetal_sex interaction for moderated effect
df$Y_adj <- df$SRS - (df$Z_hat - mean(df$Z_hat))
pgc_model <- lm(Y_adj ~ PFAS * fetal_sex, data = df)
summary(pgc_model)

pgc_fn <- function(data, indices) {
  d <- data[indices, ]
  
  # Stage 1: bridge model with SRS × sex (optional)
  bridge <- lm(m_meth ~ expr + SRS * fetal_sex, data = d)
  Z_hat <- predict(bridge)
  Y_adj <- d$SRS - (Z_hat - mean(Z_hat))
  
  # Stage 2: PGC with PFAS × fetal_sex
  pgc_fit <- lm(Y_adj ~ PFAS * fetal_sex, data = cbind(d, Y_adj))
  
  # Return main effect of PFAS (for females) only
  return(coef(pgc_fit)["PFAS"])
}

pgc_est_female <- coef(pgc_model)["PFAS"]
pgc_est_male <- coef(pgc_model)["PFAS"] + coef(pgc_model)["PFAS:fetal_sex"]

pgc_fn <- function(data, indices) {
  d <- data[indices, ]
  bridge <- lm(m_meth ~ expr + SRS * fetal_sex, data = d)
  Z_hat <- predict(bridge)
  Y_adj <- d$SRS - (Z_hat - mean(Z_hat))
  pgc_fit <- lm(Y_adj ~ PFAS * fetal_sex, data = cbind(d, Y_adj))

  beta_pfas <- coef(pgc_fit)["PFAS"]
  beta_int <- coef(pgc_fit)["PFAS:fetal_sex"]
  c(female = beta_pfas, male = beta_pfas + beta_int)
}

# Bootstrap
boot_out <- boot(df, pgc_fn, R = 100)
boot.ci(boot_out, type = "perc")
```

# Parallelized Analysis Across Isoforms

```{r parallelized_analysis, warning=F}
results_list <- vector("list", length = n_isoforms)
names(results_list) <- paste0("isoform_", 1:n_isoforms)

n_boot <- 10

for (i in 1:n_isoforms) {
  iso_name <- paste0("isoform_", i)
  
  df <- cbind(metadata_df, expr = expr_df[[iso_name]])
  
  meth_avg <- meth_long %>%
    filter(isoform == iso_name) %>%
    group_by(sample) %>%
    summarise(m_meth = mean(meth), .groups = "drop")
  
  df <- merge(df, meth_avg, by = "sample")

  # --- lavaan SEM (already includes sex and interaction terms)
  model <- '
    expr ~ a1*PFAS + a2*fetal_sex + a3*PFAS:fetal_sex
    SRS ~ b1*expr + b2*fetal_sex + b3*expr:fetal_sex + c*PFAS
    ind_female := a1 * b1
    ind_male := (a1 + a3) * (b1 + b3)
    dir := c
    total_female := dir + ind_female
    total_male := dir + ind_male
  '
  
  lavaan_fit <- tryCatch({
    sem(model, data = df, se = "bootstrap", bootstrap = n_boot)
  }, error = function(e) NULL)

  # --- COCA with PFAS × sex interaction
  coca_model <- lm(m_meth ~ PFAS + SRS * fetal_sex, data = df)
  beta_A <- coef(coca_model)["PFAS"]
  beta_Y <- coef(coca_model)["SRS"]
  psi_coca <- -beta_A / beta_Y

  coca_fn <- function(data, indices) {
    d <- data[indices, ]
    fit <- lm(m_meth ~ PFAS + SRS * fetal_sex, data = d)
    -coef(fit)["PFAS"] / coef(fit)["SRS"]
  }

  coca_boot <- boot(df, coca_fn, R = n_boot)
  coca_ci <- tryCatch(boot.ci(coca_boot, type = "perc"), error = function(e) NULL)

  # --- PGC with PFAS × sex interaction in second stage and SRS × sex in bridge
  bridge_model <- lm(m_meth ~ expr + SRS * fetal_sex, data = df)
  df$Z_hat <- predict(bridge_model)
  df$Y_adj <- df$SRS - (df$Z_hat - mean(df$Z_hat))

  pgc_model <- lm(Y_adj ~ PFAS * fetal_sex, data = df)

  pgc_fn <- function(data, indices) {
    d <- data[indices, ]
    bridge <- lm(m_meth ~ expr + SRS * fetal_sex, data = d)
    Z_hat <- predict(bridge)
    Y_adj <- d$SRS - (Z_hat - mean(Z_hat))
    pgc_fit <- lm(Y_adj ~ PFAS * fetal_sex, data = cbind(d, Y_adj))
    return(coef(pgc_fit)["PFAS"])
  }

  pgc_boot <- boot(df, pgc_fn, R = n_boot)
  pgc_ci <- tryCatch(boot.ci(pgc_boot, type = "perc"), error = function(e) NULL)

  # --- Store results
  results_list[[i]] <- list(
    lavaan_fit = lavaan_fit,
    COCA = list(
      estimate = psi_coca,
      boot = coca_boot,
      ci = coca_ci
    ),
    PGC = list(
      model = pgc_model,
      boot = pgc_boot,
      ci = pgc_ci
    )
  )
}
```

# Plotting

```{r plotting_results, warning = F, message = F}
# Initialize results data.frames
sem_pval_df <- data.frame()
coca_df <- data.frame()
pgc_df <- data.frame()

iso_names <- names(results_list)

for (i in seq_along(results_list)) {
  iso <- iso_names[i]
  res <- results_list[[i]]

  # --- SEM path p-values
  fit <- res$lavaan_fit
  if (!is.null(fit)) {
    pe <- tryCatch(parameterEstimates(fit), error = function(e) NULL)
    if (!is.null(pe)) {
      selected_paths <- pe %>%
        filter(label %in% c("a1", "a3", "b1", "b3", "c",
                            "ind_female", "ind_male", "total_female", "total_male")) %>%
        select(label, pvalue)
      if (nrow(selected_paths) > 0) {
        sem_pval_df <- bind_rows(sem_pval_df,
                                 data.frame(
                                   isoform = iso,
                                   path = selected_paths$label,
                                   pval = selected_paths$pvalue
                                 ))
      }
    }
  }

  # --- COCA model: extract PFAS, fetal_sex, and interaction p-values and CI
  coca_fit <- tryCatch(lm(m_meth ~ PFAS + SRS * fetal_sex, data = df), error = function(e) NULL)
  
  if (!is.null(coca_fit)) {
    coef_table <- coef(summary(coca_fit))
    confint_table <- tryCatch(confint(coca_fit), error = function(e) NULL)
  
    pval_pfas  <- coef_table["PFAS", "Pr(>|t|)"]
    pval_sex   <- if ("fetal_sex" %in% rownames(coef_table)) coef_table["fetal_sex", "Pr(>|t|)"] else NA
    pval_int   <- if ("SRS:fetal_sex" %in% rownames(coef_table)) coef_table["SRS:fetal_sex", "Pr(>|t|)"] else NA
  
    ci_lower <- ci_upper <- NA
    if (!is.null(confint_table) && "PFAS" %in% rownames(confint_table)) {
      ci_lower <- confint_table["PFAS", 1]
      ci_upper <- confint_table["PFAS", 2]
    }
  
    signif <- !(is.na(ci_lower) || (ci_lower < 0 & ci_upper > 0))  # TRUE if CI excludes 0
  
    coca_df <- bind_rows(coca_df, data.frame(
      isoform = iso,
      pval_PFAS = pval_pfas,
      pval_fetal_sex = pval_sex,
      pval_PFAS_fetal_sex = pval_int,
      ci_lower = ci_lower,
      ci_upper = ci_upper,
      significant = signif
    ))
  }

  # --- PGC model: extract PFAS, fetal_sex, and interaction p-values
  pgc_model <- tryCatch(res$PGC$model, error = function(e) NULL)
  pgc_boot  <- tryCatch(res$PGC$boot, error = function(e) NULL)

  if (!is.null(pgc_model) && !is.null(pgc_boot)) {
    coef_pg <- coef(summary(pgc_model))

    pval_pfas <- coef_pg["PFAS", "Pr(>|t|)"]
    pval_sex  <- if ("fetal_sex" %in% rownames(coef_pg)) coef_pg["fetal_sex", "Pr(>|t|)"] else NA
    pval_int  <- if ("PFAS:fetal_sex" %in% rownames(coef_pg)) coef_pg["PFAS:fetal_sex", "Pr(>|t|)"] else NA

    boot_estimates <- pgc_boot$t
    signif <- NA
    ci_lower <- ci_upper <- NA
    if (!is.null(boot_estimates) && length(boot_estimates) > 1) {
      ci_bounds <- quantile(boot_estimates, probs = c(0.025, 0.975), na.rm = TRUE)
      ci_lower <- ci_bounds[1]
      ci_upper <- ci_bounds[2]
      signif <- !(ci_lower < 0 & ci_upper > 0)
    }

    pgc_df <- bind_rows(pgc_df, data.frame(
      isoform = iso,
      pval_PFAS = pval_pfas,
      pval_fetal_sex = pval_sex,
      pval_PFAS_fetal_sex = pval_int,
      ci_lower = ci_lower,
      ci_upper = ci_upper,
      significant = signif
    ))
  }
}
```

## SEM p-value distributions

```{r plot_sem_pvalues}
ggplot(sem_pval_df, aes(x = pval)) +
  geom_histogram(binwidth = 0.025, fill = "#4682B4", color = "black") +
  facet_wrap(~ path, scales = "free_y") +
  theme_minimal() +
  labs(title = "SEM Path p-values Across Isoforms",
       x = "p-value", y = "Frequency")
```

```{r sem_table, echo = F}
sem_table <- data.frame(
  Path = c("a1", "a3", "b1", "b3", "c", "ind_male", "ind_female", "total_male", "total_female"),
  Pattern = c(
    "Strong left skew",
    "Uniform / flat",
    "Strong left skew",
    "Uniform",
    "Strong left skew (many significant)",
    "Uniform to right-skewed",
    "U-shaped with weak left skew",
    "Strong left skew",
    "Strong left skew"
  ),
  Interpretation = c(
    "PFAS is robustly associated with imprinting disruption across many isoforms.",
    "No consistent fetal sex interaction on PFAS-to-expression path.",
    "Imprinting disruption is significantly associated with SRS in many isoforms.",
    "No consistent interaction effect between sex and expression on SRS.",
    "PFAS has a strong total effect on SRS in many isoforms, suggesting true mediation.",
    "Some male-specific mediation is detectable, but effects are diffuse.",
    "Female-specific mediation appears inconsistent or absent.",
    "Many isoforms show strong total PFAS -> SRS effects in males.",
    "Many isoforms show strong total PFAS -> SRS effects in females."
  )
)

kable(sem_table, caption = "Interpretation of SEM Path p-value Distributions")
```

## COCA & PGC p-value distributions

```{r plot_coca_pgc_pvals, warning = F}
library(dplyr)
library(tidyr)
library(ggplot2)

# Pivot COCA to long format
coca_pvals_long <- coca_df %>%
  pivot_longer(
    cols = c(pval_PFAS, pval_fetal_sex, pval_PFAS_fetal_sex),
    names_to = "Term",
    values_to = "pval"
  ) %>%
  mutate(
    Method = "COCA",
    Term = recode(Term,
                  "pval_PFAS" = "PFAS",
                  "pval_fetal_sex" = "fetal_sex",
                  "pval_PFAS_fetal_sex" = "PFAS:fetal_sex")
  )

# Pivot PGC to long format
pgc_pvals_long <- pgc_df %>%
  pivot_longer(
    cols = c(pval_PFAS, pval_fetal_sex, pval_PFAS_fetal_sex),
    names_to = "Term",
    values_to = "pval"
  ) %>%
  mutate(
    Method = "PGC",
    Term = recode(Term,
                  "pval_PFAS" = "PFAS",
                  "pval_fetal_sex" = "fetal_sex",
                  "pval_PFAS_fetal_sex" = "PFAS:fetal_sex")
  )

# Combine and filter non-missing p-values
pval_df <- bind_rows(coca_pvals_long, pgc_pvals_long) %>%
  filter(!is.na(pval))

# Plot
ggplot(pval_df, aes(x = pval)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "black") +
  facet_grid(Method ~ Term, scales = "free", drop = TRUE) +
  theme_minimal() +
  labs(
    title = "Raw p-values for COCA and PGC Tests",
    x = "p-value", y = "Number of Isoforms"
  )
```

## COCA & PGC significance

```{r plot_significance_bar}
# Significance for PFAS (from 95% CI exclusion flag)
sig_pfas <- bind_rows(
  coca_df %>% transmute(isoform, significant, Method = "COCA", Term = "PFAS"),
  pgc_df %>% transmute(isoform, significant, Method = "PGC", Term = "PFAS")
)

# Significance for fetal_sex (p < 0.05)
sig_sex <- bind_rows(
  coca_df %>%
    filter(!is.na(pval_fetal_sex)) %>%
    mutate(significant = pval_fetal_sex < 0.05,
           Method = "COCA", Term = "fetal_sex") %>%
    select(isoform, significant, Method, Term),
  pgc_df %>%
    filter(!is.na(pval_fetal_sex)) %>%
    mutate(significant = pval_fetal_sex < 0.05,
           Method = "PGC", Term = "fetal_sex") %>%
    select(isoform, significant, Method, Term)
)

# Significance for PFAS:fetal_sex interaction (p < 0.05)
sig_int <- bind_rows(
  coca_df %>%
    filter(!is.na(pval_PFAS_fetal_sex)) %>%
    mutate(significant = pval_PFAS_fetal_sex < 0.05,
           Method = "COCA", Term = "PFAS:fetal_sex") %>%
    select(isoform, significant, Method, Term),
  pgc_df %>%
    filter(!is.na(pval_PFAS_fetal_sex)) %>%
    mutate(significant = pval_PFAS_fetal_sex < 0.05,
           Method = "PGC", Term = "PFAS:fetal_sex") %>%
    select(isoform, significant, Method, Term)
)

# Combine all
sig_df <- bind_rows(sig_pfas, sig_sex, sig_int) %>%
  mutate(
    Significance = ifelse(significant, "Significant", "Not Significant"),
    Method = factor(Method),
    Term = factor(Term, levels = c("PFAS", "fetal_sex", "PFAS:fetal_sex")),
    Significance = factor(Significance)
  )

# Plot
ggplot(sig_df, aes(x = Term, fill = Significance)) +
  geom_bar(position = "dodge", color = "black", na.rm = TRUE) +
  facet_wrap(~ Method, drop = TRUE, scales = "free") +
  scale_fill_manual(values = c("gray70", "forestgreen")) +
  theme_minimal() +
  labs(
    title = "CI- or p-value-Based Significance of Modeled Effects (COCA & PGC)",
    x = "Model Term", y = "Number of Isoforms", fill = "Significance"
  )
```

```{r interpret_NC_methods, echo = F}
coca_pgc_table <- data.frame(
  Method = c("COCA", "PGC"),
  PValue_Pattern = c(
    "Nearly all p-values > 0.9 for PFAS, sex and interaction show spurious signal",
    "PFAS p-values strongly left-skewed; interaction and sex terms right-skewed"
  ),
  CI_Significance = c(
    "100% of isoforms significant for PFAS, 0% for sex or interaction",
    "~80% of isoforms significant for PFAS, 0% for sex or interaction"
  ),
  Interpretation = c(
    "COCA overestimates true mediation effects, detecting far more than the simulated 50%.",
    "PGC overestimates true mediation effects, detecting far more than the simulated 50%. Although more robust to confounding, it still inflates Type I error under partial confounding and ideal proxy assumptions."
  )
)

kable(coca_pgc_table, caption = "Summary of COCA and PGC Results for Isoform-Level Mediation with Sex Interaction")
```

# Regression adjustment models with control outcome variable as covariate

```{r, regression_adjustment_models}
# Example for one isoform
iso_idx <- 27
meth_proxy <- rowMeans(maternal_meth_list[[iso_idx]])

# Include sex main effect and interaction with expression
model <- lm(SRS ~ PFAS + expr_ratios[, iso_idx] * fetal_sex + meth_proxy)

summary(model)
```

```{r, loop_rgms_across_isoforms}
# Initialize results table for moderated direct adjustment
adjusted_lm_results <- data.frame(
  isoform = 1:n_isoforms,
  coef_expr = NA,           p_expr = NA,           sig_ci_expr = NA,
  coef_sex = NA,            p_sex = NA,            sig_ci_sex = NA,
  coef_interaction = NA,    p_interaction = NA,    sig_ci_interaction = NA
)

for (i in 1:n_isoforms) {
  meth_proxy <- rowMeans(maternal_meth_list[[i]])
  fit <- lm(SRS ~ PFAS + expr_ratios[, i] * fetal_sex + meth_proxy)
  confint_fit <- confint(fit)
  coef_names <- names(coef(fit))

  expr_term <- "expr_ratios[, i]"
  sex_term  <- "fetal_sex"
  int_term  <- "expr_ratios[, i]:fetal_sex"

  # Expression term
  if (expr_term %in% coef_names) {
    idx <- which(coef_names == expr_term)
    adjusted_lm_results$coef_expr[i] <- coef(fit)[idx]
    adjusted_lm_results$p_expr[i] <- summary(fit)$coefficients[idx, "Pr(>|t|)"]
    adjusted_lm_results$sig_ci_expr[i] <- !(0 >= confint_fit[idx, 1] & 0 <= confint_fit[idx, 2])
  }

  # Fetal sex term
  if (sex_term %in% coef_names) {
    idx <- which(coef_names == sex_term)
    adjusted_lm_results$coef_sex[i] <- coef(fit)[idx]
    adjusted_lm_results$p_sex[i] <- summary(fit)$coefficients[idx, "Pr(>|t|)"]
    adjusted_lm_results$sig_ci_sex[i] <- !(0 >= confint_fit[idx, 1] & 0 <= confint_fit[idx, 2])
  }

  # Interaction term
  if (int_term %in% coef_names) {
    idx <- which(coef_names == int_term)
    adjusted_lm_results$coef_interaction[i] <- coef(fit)[idx]
    adjusted_lm_results$p_interaction[i] <- summary(fit)$coefficients[idx, "Pr(>|t|)"]
    adjusted_lm_results$sig_ci_interaction[i] <- !(0 >= confint_fit[idx, 1] & 0 <= confint_fit[idx, 2])
  }
}

# Stack into long format for p-values
pval_df <- bind_rows(
  adjusted_lm_results %>% select(isoform, pval = p_expr) %>% mutate(Term = "expr"),
  adjusted_lm_results %>% select(isoform, pval = p_sex) %>% mutate(Term = "fetal_sex"),
  adjusted_lm_results %>% select(isoform, pval = p_interaction) %>% mutate(Term = "expr:fetal_sex")
)

# Add method label
pval_df$Method <- "DirectAdj"

ggplot(pval_df, aes(x = pval)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "black") +
  facet_grid(Method ~ Term, scales = "free", drop = TRUE) +
  theme_minimal() +
  labs(
    title = "Raw p-values for Direct Adjustment Tests",
    x = "p-value", y = "Number of Isoforms"
  )

sig_df <- bind_rows(
  adjusted_lm_results %>% transmute(isoform, Term = "expr",         sig = sig_ci_expr),
  adjusted_lm_results %>% transmute(isoform, Term = "fetal_sex",    sig = sig_ci_sex),
  adjusted_lm_results %>% transmute(isoform, Term = "expr:fetal_sex", sig = sig_ci_interaction)
) %>%
  mutate(
    Method = "DirectAdj",
    Significance = ifelse(sig, "Significant", "Not Significant")
  )

ggplot(sig_df, aes(x = Term, fill = Significance)) +
  geom_bar(position = "dodge", color = "black") +
  facet_wrap(~ Method) +
  scale_fill_manual(values = c("gray70","forestgreen")) +
  theme_minimal() +
  labs(
    title = "CI-Based Significance of Direct Adjustment Terms",
    x = "Model Term", y = "Number of Isoforms", fill = "Significance"
  )
```

```{r interpret_regression, echo=F}
direct_adj_table <- data.frame(
  Method = "Direct Adjustment (Huang, 2021)",
  PValue_Pattern = "Uniform (no enrichment near 0)",
  CI_Significance = "Few isoforms significant",
  Interpretation = "Adjusting directly for methylation removes most expression–SRS signal, suggesting either methylation absorbs the effect (overadjustment), or residual confounding remains. Consistent with Huang (2021), this approach is conservative but stable under real-data-like assumptions."
)

kable(direct_adj_table, caption = "Summary of Direct Adjustment Results (Huang, 2021 Approach)")
```


# Calculate Type I Error Rates

```{r calc_type1_error}
# For a null simulation, we expect these rates to be close to 0.05 (your alpha level)
# for true null effects, indicating correct Type I error control.
# Rates significantly above 0.05 suggest inflated Type I error.

type1_error_results <- list()

# 1. SEM Type I Error
# Focusing on the mediation paths (a1, b1) and indirect effects (ind_female, ind_male)
# which are set to null in the simulation scenario.
sem_type1 <- sem_pval_df %>%
  filter(path %in% c("a1", "b1", "ind_female", "ind_male")) %>%
  group_by(path) %>%
  summarise(
    Type1_Error_Rate = mean(pval < 0.05, na.rm = TRUE),
    .groups = "drop"
  )
type1_error_results$SEM <- sem_type1

# 2. COCA Type I Error
# For COCA, 'significant' is determined by whether the 95% CI for the PFAS effect excludes 0.
# For other terms (fetal_sex, SRS:fetal_sex interaction), we use a p-value threshold.
coca_type1 <- coca_df %>%
  summarise(
    Type1_Error_Rate_PFAS = mean(significant, na.rm = TRUE), # Uses the 'significant' flag from CI
    Type1_Error_Rate_fetal_sex = mean(pval_fetal_sex < 0.05, na.rm = TRUE),
    Type1_Error_Rate_SRS_fetal_sex_int = mean(pval_PFAS_fetal_sex < 0.05, na.rm = TRUE) # This is for the SRS:fetal_sex interaction in the COCA model
  ) %>%
  pivot_longer(everything(), names_to = "path", values_to = "Type1_Error_Rate") %>%
  mutate(path = gsub("Type1_Error_Rate_", "", path))
type1_error_results$COCA <- coca_type1

# 3. PGC Type I Error
# For PGC, 'significant' for PFAS is also based on CI exclusion of 0.
# For other terms (fetal_sex, PFAS:fetal_sex interaction), p-value threshold is used.
pgc_type1 <- pgc_df %>%
  summarise(
    Type1_Error_Rate_PFAS = mean(significant, na.rm = TRUE), # Uses the 'significant' flag from CI
    Type1_Error_Rate_fetal_sex = mean(pval_fetal_sex < 0.05, na.rm = TRUE),
    Type1_Error_Rate_PFAS_fetal_sex_int = mean(pval_PFAS_fetal_sex < 0.05, na.rm = TRUE) # This is for the PFAS:fetal_sex interaction in the PGC model
  ) %>%
  pivot_longer(everything(), names_to = "path", values_to = "Type1_Error_Rate") %>%
  mutate(path = gsub("Type1_Error_Rate_", "", path))
type1_error_results$PGC <- pgc_type1

# 4. Regression Adjustment Type I Error
# 'sig_ci_*' flags are based on CI excluding 0.
adj_type1 <- adjusted_lm_results %>%
  summarise(
    Type1_Error_Rate_expr = mean(sig_ci_expr, na.rm = TRUE),
    Type1_Error_Rate_sex = mean(sig_ci_sex, na.rm = TRUE),
    Type1_Error_Rate_expr_fetal_sex_int = mean(sig_ci_interaction, na.rm = TRUE) # This is for the expr:fetal_sex interaction in the direct adjustment model
  ) %>%
  pivot_longer(everything(), names_to = "path", values_to = "Type1_Error_Rate") %>%
  mutate(path = gsub("Type1_Error_Rate_", "", path))
type1_error_results$DirectAdj <- adj_type1
```

```{r print_type1_error, echo=F}
# Combine all tibbles into a single data frame
combined_type1_errors <- bind_rows(
  type1_error_results$SEM %>% mutate(Method = "SEM"),
  type1_error_results$COCA %>% mutate(Method = "COCA"),
  type1_error_results$PGC %>% mutate(Method = "PGC"),
  type1_error_results$DirectAdj %>% mutate(Method = "Direct Adjustment")
) %>%
  # Optionally reorder columns for better presentation
  select(Method, path, Type1_Error_Rate) %>%
  # Optionally rename paths for clarity if desired (e.g., "a1" to "PFAS -> Expression")
  mutate(
    path = case_when(
      path == "a1" ~ "PFAS -> Expression (a1)",
      path == "b1" ~ "Expression -> SRS (b1)",
      path == "ind_female" ~ "Indirect Effect (Female)",
      path == "ind_male" ~ "Indirect Effect (Male)",
      path == "PFAS" ~ "PFAS Main Effect",
      path == "fetal_sex" ~ "Fetal Sex Main Effect",
      path == "SRS_fetal_sex_int" ~ "SRS:Fetal Sex Interaction",
      path == "PFAS_fetal_sex_int" ~ "PFAS:Fetal Sex Interaction",
      path == "expr" ~ "Expression Main Effect",
      path == "sex" ~ "Sex Main Effect",
      path == "expr_fetal_sex_int" ~ "Expression:Fetal Sex Interaction",
      TRUE ~ path # Keep others as is
    )
  )

# Print the combined table using kable
kable(combined_type1_errors, 
      caption = "Type I Error Rates Across Different Methods and Causal Paths (Null Simulation)",
      digits = 3) # Format numbers to 3 decimal places for readability
```

#### Interpretation of Overall Type 1 Error Rates by Method

##### SEM

SEM shows significant inflation of Type I error for the direct mediation paths (a1 and b1), which are the building blocks of the mediation. The indirect effects, while still showing some inflation, are closer to the nominal level. This suggests that the underlying model assumptions or the sample size might be contributing to issues with accurately estimating and testing these specific direct paths under the null.

##### COCA

COCA exhibits an extremely high Type I error rate for the primary effect it estimates (PFAS), making its results highly unreliable under the null. This warrants a close examination of the COCA implementation or its underlying assumptions for this particular simulation.

##### PGC

PGC, similar to COCA, severely inflates the Type I error for the PFAS main effect, raising concerns about its reliability in detecting null effects in this simulation context.

##### Direct Adjustment

The direct adjustment method demonstrates relatively better control of Type I error compared to SEM, COCA, and PGC for the effects it estimates. While there's a slight inflation for the expression main effect, it's much closer to the nominal alpha level. This suggests it might be a more conservative but reliable approach under the simulated conditions for Type I error control.